High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-09-29 manuscript/Manuscript.tex

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers

@@

@@

This proposed magmatic evolution of the Duluth Complex is in distinct contrast to the emplacement of the Beaver River diabase which rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \citeA{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported into shallow magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Fig. \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite    interstitial to the plagioclase are typical of Beaver River anorthosites (Fig, \ref{}).

@@

% \begin{figure}
% \noindent\includegraphics[width=\textwidth]{}
% \caption{\small{}
% \label{fig:Petro_QDM}
% \end{figure}

@@

In this study, we conduct rock magnetic experiments with a purpose of gaining magnetic mineralogy insight into the paleointensity results of the anorthosite and diabase. Backfield curves were measured at room temperature using a Micromag Princeton Measurements vibrating sample magnetometer (VSM) with nominal sensitivity of $5 \times 10^{-9}$ Am\textsuperscript{2}, and a Lake Shore 8600 series VSM with nominal sensitivity of xxxxxxxxxx. The calculated coercivity spectra were subsequently decomposed into one ore more components using skew-normal distributions following the method of \citeA{Maxbauer2016a}. The fitted components are then used to interpret the characteristics of different populations of magnetic particles within specimens. We also used the magnetic property measurement system (MPMS) at the IRM to aid in the identification of magnetic mineral. The operational sequence follows an in-field cooling step in a 2.5 T field from room temperature to 10 K, a zero-field cooling from room temperature to 10 K, and a final cooling from room temperature to 10 K after an IRM acquisition at room temperature. The magnetic moment transitions at critical temperatures revealed through MPMS experiments are used to identify magnetic minerals such as magnetite \cite{Verwey1939a} within specimens. 
% it would actually be nice to show the MPMS data (although it could be a bit redundant). The MPMS data nicely shows the BD and failed AX have the Verwey transition suppressed to lower temperatures and the successful AX have Verwey transition close to 120 K - the typical value for stochiometric magnetite. I don't think I will be using the hysteresis loops.

@@

\subsection*{Rock magnetism}
Rock magnetic data support the interpretation that the successful anorthosite specimens have dominant magnetic carriers with magnetic properties similar to stochiometric, non-interactive, single domain magnetite, whereas those anorthosites that failed the paleointensity result selection and all diabase have more pronounced non-ideal carriers. MPMS data show sister specimens to paleointensity specimens typically have dominant magnetic mineralogy of magnetite as can be evidenced through the Verwey transitions (Figure \ref{fig:MPMS}; \citeA{Verwey1939a}). However, specimens from sites that yield successful paleointensity results display Verwey transition near 120 Kelvin (K) as expected for stochiometric low-Ti magnetite (\cite{Ozdemir1993a}). On the other hand, the diabase specimens and anorthosite specimens that did not yielded accepted paleointensity results typically have suppressed Verwey transition toward lower temperatures, indicating that the titanomagnetite within the specimens likely experienced variable degrees of oxidation \cite{Ozdemir1993a}.

Specimen coercivity spectra further support that our paleointensity selection criteria are successful in selecting experiment results. A compilation of median destructive field (MDF) in Figure \ref{fig:coercivity} based on all of our backfield experiments shows that successful anorthosites have distinctly higher average MDF range than others. Example single component skew-normal distribution fits of coercivity spectra based on backfield experiments on specimen chips sister to paleointensity specimens show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity distributions around 80 mT,
very similar to that of a bulk sample composed of no-interactive single domain magnetite (Figure \ref{fig:coercivity}). In contrast, specimens from anorthosite that did not pass paleointensity selection, as well as diabase specimens, tend to have coercivity distributions over a significantly lower range ($\sim$30 mT). This is consistent with the interpretation that they have a more dominant population of high Ti content, multi-domain grains or vortex state grains.

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{MPMS.pdf}
\centering
\caption{\footnotesize{}}
\label{fig:MPMS}
\end{figure}

@@

From the thermal history model we can also estimate the time duration which the diabase and anorthosite spent to block in the primary component of their remanence. We found the cooling time to be on the order of magnitude of $\sim$1 kyr from 580 to 500 $^\circ$C, which corresponds to a cooling rate of $1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at about $1.3\times10^{-1}$ $^\circ$C s$^{-1}$ if we assume it takes 10 minutes in lab condition to cool through the same temperature interval. As a result, this significant cooling rate difference leads to a prediction of 35.5\% overestimate of true field \cite{Halgedahl1980a}. We therefore correct our paleointensity results by a factor of 0.74.

\subsection*{Late Mesoproterozoic geomagnetic field and the Precambrian observation records}
Figure \ref{fig:PINT_plot} shows all accepted anorthosite and diabase specimen paleointensity results and their site-level means with one standard deviation envelopes. An anorthosite site mean paleointensity weighted by site standard deviations is plotted as a dashed orange line, and the standard deviation envelope as dashed green lines. The anorthosites are also grouped by their geographical vicinities \ref{fig:PINT_plot}. The site-level results are shown in Table \ref{tab:PINT_result}. Anorthosite sites AX10, 11, 12, 13, 16 yielded very high within-site coherence (Figure \ref{fig:Arai_plots, fig:PINT_plot}), which suggest high-confidence site-level paleointensity estimates. Clearly there are two populations of paleointensity results: a group of high estimates from anorthosite site 11 and 13 which have very high within-site consistency and are in agreement with the results from BRD site 2; and another group of low estimates from anorthosite site 6, 8, 10, 12, and 16, which also have tight agreement across sites.

All units have Q$_{PI}$ scores \cite{Biggin2014a} ranging between 6 and 7 (Table \ref{tab:QPI}). All sites passed AGE, TRM and ALT (pTRM checks passed and high-temperature steps that looked to be suffering from alteration were excluded), ACN and MD (no sig- nificant sagging or zigzagging).

% need to export the figure with tight boundaries
\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\small{ }}
\label{fig:PINT_cooling_corrected}
\end{figure}

>>>>

\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}

\linenumbers

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from Duluth Complex Anorthositic Series (left column) and anorthosite xenolith MS99033 from the Beaver River diabase (right column). Petrographic images (A-D) show Duluth Complex anorthosite have euhedral, elongate plagioclase crystals with interlocking texture and strong igneous foliation. The plagioclase can contain abundant Fe-Ti oxide needles (C) that have preferred orientations that are often parallel with the (010) planes of plagioclase. In contrast, plagioclase of the anorthosite xenolith MS99033 do not enclose oxide needles. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the oxide needles within Duluth plagioclase are strongly magnetic as evidenced through the overlapping magnetic dipole fields exerted above the oxides in (E). In comparison, the plagioclase of MS99033 is not magnetic. }}
\label{fig:Petro_QDM}
\end{figure}

@@

This proposed magmatic evolution of the Duluth Complex is in distinct contrast to the emplacement of the Beaver River diabase which rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \citeA{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported into shallow magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Fig. \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite cinterstitial to the plagioclase are typical of Beaver River anorthosites (Fig, \ref{}).

@@

@@

In this study, we conduct rock magnetic experiments with a purpose of gaining magnetic mineralogy insight into the paleointensity results of the anorthosite and diabase. Backfield curves were measured at room temperature using a Micromag Princeton Measurements vibrating sample magnetometer (VSM) with nominal sensitivity of $5 \times 10^{-9}$ Am\textsuperscript{2}, and a Lake Shore 8600 series VSM with nominal sensitivity of xxxxxxxxxx. The calculated coercivity spectra were subsequently decomposed into one ore more components using skew-normal distributions following the method of \citeA{Maxbauer2016a}. The fitted components are then used to interpret the characteristics of different populations of magnetic particles within specimens. We also used the magnetic property measurement system (MPMS) at the IRM to aid in the identification of magnetic mineral. In the field-cooled (FC) experiments shown in Figure \ref{fig:MPMS}, the magnetization was measured upon warming following the specimen having cooled in an applied field of 2.5 T from 300 to 10 K. In the zero-field-cooled (ZFC) experiment, a low-temperature saturation isothermal remanence (LTSIRM) of 2.5 T was applied at 10 K after the specimen cooled in a (near-)zero field. In the room-temperature saturation isothermal remanence (RTSIRM) experiment, the sample was pulsed with a 2.5 T field at room temperature ($\sim$300 K) and then cooled to 10 K and warmed back to room temperature in a (near)zero field. The magnetic moment transitions at critical temperatures revealed through MPMS experiments are used to identify magnetic minerals such as magnetite \cite{Verwey1939a} within specimens. 
% it would actually be nice to show the MPMS data (although it could be a bit redundant). The MPMS data nicely shows the BD and failed AX have the Verwey transition suppressed to lower temperatures and the successful AX have Verwey transition close to 120 K - the typical value for stochiometric magnetite. I don't think I will be using the hysteresis loops.

@@

\begin{table}[]
\caption{\footnotesize{Summary statistics for the Q$_{PI}$ quality criteria of \citeA{Biggin2014a}.}}
\centering
\begin{tabular}{ccccccccccccc}
\hline
Site & N  & Age (Ma) & Method & AGE & STAT & TRM & ALT & MD & ACN & TECH & LITH & QPI \\ \hline
AX6  & 3  & 1091.8   & T+     & 1   & 0    & 1   & 1   & 1  & 1   & 0    & 0    & 5   \\
AX8  & 3  & 1091.8   & T+     & 1   & 0    & 1   & 1   & 1  & 1   & 0    & 0    & 5   \\
AX10 & 3  & 1091.8   & T+     & 1   & 0    & 1   & 1   & 1  & 1   & 0    & 0    & 5   \\
AX12 & 6  & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\
AX16 & 11 & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\
AX11 & 7  & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\
AX13 & 7  & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\ \hline
BD2  & 9  & 1091.7   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\ \hline
\end{tabular}
\label{tab:QPI}
\end{table}

\subsection*{Rock magnetism}
Rock magnetic data support the interpretation that the successful anorthosite specimens have dominant magnetic carriers with magnetic properties similar to stochiometric, non-interactive, single domain magnetite, whereas those anorthosites that failed the paleointensity result selection and all diabase have more pronounced non-ideal carriers. MPMS data show sister specimens to paleointensity specimens typically have dominant magnetic mineralogy of magnetite as can be evidenced through the Verwey transitions (Figure \ref{fig:MPMS}; \citeA{Verwey1939a}). However, specimens from sites that yield successful paleointensity results display Verwey transition near 120 Kelvin (K) as expected for stochiometric low-Ti magnetite (\cite{Ozdemir1993a}). On the other hand, the diabase specimens and anorthosite specimens that did not yielded accepted paleointensity results typically have suppressed Verwey transition toward lower temperatures (Figure \ref{fig:MPMS}), indicating that the specimens likely have a ferromagnetic mineralogy with relatively high Ti content and/or with variable degrees of oxidation \cite{Ozdemir1993a}.

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{MPMS.pdf}
\centering
\caption{\footnotesize{Low-temperature magnetic property measurement system (MPMS) experiment results. In the field-cooled (FC) experiments, the magnetization was measured upon warming following the specimen having cooled in an applied field of 2.5 T from 300 to 10 K. In the zero-field-cooled (ZFC) experiment, a low-temperature saturation isothermal remanence (LTSIRM) of 2.5 T was applied at 10 K after the specimen cooled in a (near-)zero field. In the room-temperature saturation isothermal remanence (RTSIRM) experiment, the sample was pulsed with a 2.5 T field at room temperature ($\sim$300 K) and then cooled to 10 K and warmed back to room temperature in a (near)zero field. Specimen AX6-1r is from anorthosite AX6 which passed our paleointensity selection. It has a well-defined Verwey transition $\sim$120 K. Specimens AX3-2r-ni and BD2-2r have prominent Verwey transition but the transition temperatures are suppressed below $\sim$120 K. }}
\label{fig:MPMS}
\end{figure}

Specimen coercivity spectra further support that our paleointensity selection criteria are successful in selecting experiment results. A compilation of median destructive field (MDF) in Figure \ref{fig:coercivity} based on all of our backfield experiments shows that successful anorthosites have distinctly higher average MDF range than others. Example single component skew-normal distribution fits of coercivity spectra based on backfield experiments on specimen chips sister to paleointensity specimens show that anorthosites that yielded successful paleointensity results can have magnetic grain populations with peak coercivity distributions around 80 mT, very similar to that of a bulk sample composed of no-interactive single domain magnetite (Figure \ref{fig:coercivity}). In contrast, specimens from anorthosite that did not pass paleointensity selection, as well as diabase specimens, tend to have coercivity distributions over a significantly lower range ($\sim$30 mT). This is consistent with the interpretation that they lack a dominance of single domain magnetite grains in their ferromagnetic mineral assemblages.

@@

From the thermal history model we can also estimate the duration which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully block in their characteristic natural remanence component ($\sim$500\textdegree C; \citeA{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at about $1.3\times10^{-1}$ $^\circ$C s$^{-1}$ if we assume it takes 10 minutes in lab condition to cool through the same temperature interval. As a result, this significant cooling rate difference leads to a prediction of 35.5\% overestimate of true ancient field \cite{Halgedahl1980a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate corrected specimen paleointensity estimates together with specimen- and site-level weighted means are shown in Figure \ref{fig:PINT_cooling_corrected}. Taking all accepted anorthosite paleointensity results, they have a site mean absolute paleointensity estimates of 47.90 $\pm$ 9.32 $\mu$T (Figure \ref{fig:PINT_cooling_corrected}).

\subsection*{Late Mesoproterozoic geomagnetic field and the Precambrian observation records}
Figure \ref{fig:PINT_plot} shows all accepted anorthosite and diabase specimen paleointensity results and their site-level means with one standard deviation envelopes. An anorthosite site mean paleointensity weighted by site standard deviations is plotted as a dashed orange line, and the standard deviation envelope as dashed green lines. The anorthosites are also grouped by their geographical vicinity \ref{fig:PINT_plot}. The site-level results are shown in Table \ref{tab:PINT_result}. Anorthosite sites AX10, 11, 12, 13, 16 yielded very high within-site coherence (Figure \ref{fig:Arai_plots, fig:PINT_plot}), which suggest high-confidence site-level paleointensity estimates. Clearly there are two populations of paleointensity results: a group of high estimates from anorthosite site 11 and 13 which have very high within-site consistency and are in agreement with the results from BRD site 2; and another group of low estimates from anorthosite site 6, 8, 10, 12, and 16, which also have tight agreement across sites.

All units have Q$_{PI}$ scores \cite{Biggin2014a} ranging between 5 and 6 (Table \ref{tab:QPI}). All sites passed AGE, TRM and ALT (pTRM checks passed and high-temperature steps that looked to be suffering from alteration were excluded), ACN and MD (no significant sagging or zigzagging).

% need to export the figure with tight boundaries
\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\small{Summary plot of cooling rate corrected specimen absolute paleointensity results (square and diamond symbols) from this study and their weighted averages at site (grey bar) and locality level (orange bar). }}
\label{fig:PINT_cooling_corrected}
\end{figure}

considering \citeA{Nagy2021a} developed a new cooling rate correction model for uniaxial single domain magnetite and suggest that the model of \citeA{Halgedahl1980a} tend to over-correct for cooling rates and lead to underestimating ancient fields. If our absolute paleointensity values are an underestimating,
